NATIONALIST AND HOME FIELD BIAS IN SPORT PERFORMANCE EVALUATIONS

### 2019-07-25 revised-manuscript/response_to_reviewers.Rmd

---
title: "Response to EJSS reviewers' comments on manuscript #2019-0005"
output:
  pdf_document:
    keep_tex: yes
    latex_engine: xelatex
    template: /home/torkildl/Dropbox/Setups/thl/rmd-xelatex-template-memo.tex
  html_document: default
bibliography: /home/torkildl/Dropbox/Literature/thl-main-library.bib
csl: /home/torkildl/Dropbox/Literature/csl-files/american-sociological-review.csl
anonymous: yes
fontsize: 12pt
geometry: margin=2cm
header-includes:
    - \usepackage{placeins}
    - \usepackage{dcolumn}
    - \usepackage{tabularx}
    - \usepackage{lscape}
    - \usepackage{caption}
fig_caption: false
fig.align: "center"
fig.width: 5
---

We are grateful for being granted the opportunity to revise and resubmit our manuscript. We have considered all the comments by both the reviewers and the Editor. They have helped us craft a stronger manuscript than we otherwise would be able to do. We believe our responses to the comments and criticisms made have made the work a stronger contribution to the literature.

In what follows, we respond to each of the comments and explain how we accordingly have adapted our work. In the cases where we have chosen not to follow the advice, we motivate our choices for doing so.

Thank you for reading and commenting on our work!

# Responses to comments made by the Editor

From the document attached with the Editor's Decision letter, we identified twelve comments and criticisms. We list the items as we have read them below, in order of appearance in the document, along with a brief statement on how we have dealt with the issue identified by the Editor.

>*1. Overemphasis on the prestige of expert judgement and being a sports judge.*

We have reworded the period on prestige of sport judges.

>*2. The size of the scholarly literature*

We have rephrased the first period of the literature review.

>*3. Typo: Em-dash before "effectively handing them the gold medal"*

The period is rewritten in the new version.

>*4. Typo: "Our second *hypotheses*" should be "Our second hypothesis"*

This is corrected in the new version.

>*5. The age of the data used in the analysis.*

In the new version, we also use data from the 2015-2016 season's World Cup competitions. We describe how we obtained these data and how they were treated in detail under our responses to reviewer \#2's comments.

>*6. The argument that results from ski jumping competitions may have relevance beyond the sport, needs a more solid foundation and a more precise characterization of ski jumping.*

We have toned down this argument somewhat in the new version.

>*7. The Editor appreciates the thorough explanation of statistical models used.*

Thank you! In this version, we have only changed small bits of material to make the models even easier to understand.

>*8. Sources of national differences in tendencies of national bias (french vs. others) -- what may they be? The Editor suggests that the potential growth in the sport nationally may be a factor.*

We now build upon sociological theory on evaluation to propose hypotheses on variation in biases in the field of ski jumping.

>*9. Our estimates of bias are less than Zitzewitz', which may be the result of a general professionalization of the sport. New data may cast further light on this issue.*

We now, with our new data, explicitly discuss and test a hypothesis of change over time.

>*10. Editor suggest that safeguards against biases are already in place (5 judges from different nations; lowest and highest scores excluded).*

This is of course correct. In the new version, we make more room for discussing these safgeguards and their effects.

>*11. Editor suggests reference to sports-related political turmoil is too far fetched. Instead, we should provide more social context of the ski jumping sport.*

We have followed this advice, and removed the political turmoil-related reference.

>*12. Reference list: There were missing data for two references (Sampaio 2012 and Yang 2006).*

The appropriate information has been added for these references.

\smallskip

We are grateful to the Editor for her/his comments on our manuscript. They helped us writerethink and rewrite the work in a way that we believe is a stronger contribution to the field. Thank you!

# Responses to comments made by reviewer \#2

Let us first express our thanks for the Reviewer's thoughtful comments on our manusript. In the folling paragraphs we respond to each of the comments and criticisms. We quote the Reviewer's comments (with minor edits) first before offering our response.

> *Distortion effects on the evaluation of performance in ski jumping are to be investigated. One is the national origin of the  judges and the other is the home advantage of the athletes. The research situation on the problem of the influence of nationality seems to be quite well worked out. The following hypotheses are formulated (p. 6):*

> *1) There will be a "positive distortion" of the results when the judges evaluate performances of athletes who come from the same nation as themselves.*

> *2) Another hypothesis claims that it is precisely the assumed effect that diminishes in strength when other variables are controlled (here I wonder about the sense). It is normal and plausible that the more variables are put into a model, the lower the relative explanatory share of a single variable. It would be interesting if the explanatory value of a variable remained high despite the control of disturbing variables (secondary variance).*

Yes, it is typical that when a new variable (Z) is introduced the regression coefficient of a first variable (X) may change both magnitude and sign if the new variable is correlated with X. In our case, it may that competitors from certain nations are better on average than competitors from other nations, and it may be that the above-average nations also are more often represented on judge panels. This will induce a correlation between the same-nation indicator (X) and qualities of the performance (Z). Controlling for Z, should then reduce the magnitude of the regression coeffient for X.

Our estimate from model 1, the "naïve" model, may be overestimating the strength of bias for such reasons, and in model 2 we control for *all* unobserved characteristics of the jump that affects the judge scores.

> *3) Finally, in a third hypothesis, cultural influences on stylistic elements are assumed (and I wonder whether this would not be an explanation for the variance of the variable "national influence"). So why is this separated? The authors differentiate between nations with a strong tradition of ski jumping and those with a rather weak tradition. In a first step there are some points critical to mention:*

> *1) The authors use a fixed effects-term for each jump in their regression-model to remove confounding factors and thereby estimate national bias more precisely. This is done under the assumption that the other - in this case four - judges give an unbiased score. Although this often so called „true score" is the common modus operandi in terms of the questions to be answered, the authors are to be pointed out on the fact that this assumption only holds true if no set of five judges contains two or more judges of identical nationality. Specific rules from the regarded years concerning the composition of the judging panel in ski jumping could be presented to support the assumption in that sense.*

Yes, we use a fixed-effects term for each jump to remove confounding factors from characteristics of the jump. However, we believe this model does _not_ rely on an assumption of unbiasedness of the other four scores nor an assumption that there is only one judge from the competitor's nation in the judge panel. If there are multiple judges from the competitor's country, there will simply be multiple 1's among the five observations that make up the observations of judge scores for each jump. What identifies the B (in the revised ms denoted $\beta$) parameter in the models, is the within-jump *variation* in judge-competitor correspondence. The identification criterion is that the data must include  some performances (jumps) where there is one (or more) judges from the same nation as the competitor, and thus scores 1 on the dummy variable for national bias. As we have quite a few of such performances, we are able to identify and estimate an effect of nationalistic bias on judge scores.

> *2) Additionally the first part oft the third hypothesis proposed by the authors seems not to be testable the way it is formulated, it is rather formulated as an additional assumption.  („A third hypothesis is that this bias may be partly due to cultural influences on stylistic elements,[…]"). For the second part of the third hypothesis however, there was no test of the hypothesis performed, nor was the concept of differing countries according to tradition for ski jumping further elucidated.*

We agree with you that the third hypothesis was mangled and untestable the way it was formulated. In the revised version, we have rewritten the hypothesis section and motivated the hypotheses better, in accordance with the new theoretical background.

> *3) Furthermore on page 12 it is stated that there would be a fluctuation in the number of observations due to missing values in the independent variables. Table 2 yet shows 18.860 observations for all three models presented, which indicates that all the available observations (3772 jumps x 5 judges) have been used to calculate the first three estimations.*

Thank you for pointing this out. This was an error on our part. It is corrected in the current version of the manuscript.

> *Home advantage is divided into a physical effect and a social effect. These constructs should be explained more clearly (see e.g. p. 10). I suspect that - according to the state of research (referenced on p. 5) - actual factors influencing performance exist (the sporting performance in the home country is actually on average higher than in a host country), but also influenced due to social pressure from the audience on the adjudicators, which can be assessed as higher in the home countries of the athletes than in the host countries (this is also explained on p. 10, this would then again be a distortion of judgement).*

We have expanded our discussion of home field advantages somewhat in this version, and make amongst other things the discintion you point to in your comment. 
TODO: OR SHOULD WE JUST DROP HOME FIELD?

> *It is expectable that the more variables are put into a model, the lower the relative explanatory share of a single variable. It would be interesting if the explanatory value of a variable remained high despite the control of disturbing variables (secondary variance).*

> *Some remarks on the regression model: The following regression models are proposed.*

>*National affiliation of the judges:*
>* sijp = B-(Lj- ϕ(I = J)) + qp + ϵijp*
>* s = score*
>* B = coefficient*
>* Φ = Dummy variable (nationality judge and athlete equal or unequal)*
>* L = leniency ("mildness" of the judges)*
>* q = jump fixed effect (as far as I understood it is the "ski jump tradition" of a country, is also explained 
>* as dummy  variable, probably nations with ski jump tradition and those without such a tradition??, that 
>* should be explained better)*
>* ϵ = disturbance term*
>* (Unclear: where is the actual power to be measured)?*

As the Editor seemed to be satisfied with the explanation of the statistical models, we have only made smaller refinements to the text on this area. We are willing, of course, to provide even deeper explanations than we already give, if need be.

A brief comment: The "jump fixed effect" represents a dummy variable for each jump (each performance), which means that we "trend" the judge scores against their mean. This implies that the model is the "within-jump" estimator of nationalistic biases.

<!---
>* Home advantage (physical):*
>* s'irs = B'-( I = H) + ai + r + ϵijp*
>* (shouldn't it be ϕ'(I = H)? Then ϕ' would be a dummy variable for the match between the venue and the athlete's country of origin.)*
>* s' = points for the width (so no posture note; in the text on p. 11 the ' is missing at s!!,)*
>* a = "seasonal effect"?*
>* r = nature of the hill?*
>* Home advantage (social portion):*
>* sijprs = B-(I = H) + ais + r + lj + xp + ϵijp*
>* supplementary (to previous)*
>* s = again posture mark*
>* l = mildness of the judges*
>* x = should represent jump characteristics (width, speed ??)*

--->

>* Formal notes: The following sources do not appear in the bibliography: Meyer & Booker, 1991; Bursell 2012, both indicated on p. 2, are not listed in the bibliography.*

This has been corrected in the revised version.

>* Typos on p. 5 Line 5: pscyhological => psychological*

This is corrected in the revised version.

>* Overall I recommend major revisions. In general: The limitations of the study should be explained better.*

TODO: In the revised version, we have tried to make the limitations of our study clearer. To this end, we have added ... ...

Thank you for taking your time to read and commenting upon on our work. In our view that your comments made us produce a stronger manuscript than the original submission, and we hope you concur.

# Response to comments made by reviewer \#3

We respond the comments made by reviewer \#3 below.

>* The manuscript is a quantitative analysis of judges' biases and home field advantages on results in ski jumping, based on data from the FIS World Cup in 2006-2008. In general, the subject is relevant and interesting; and authors apply sound methodology and statistics. All formal academic standards are met, and the paper is well written. However, data are rather old, even though they could be easily complemented by more recent data.*

In this version, we use both our original data from the mid 2000s, but also a new data set. This data set was created by scraping the FIS website using automated procedures. In short, we wrote pieces of software that 
1. walked through the FIS website and identified URLs pointing to results in PDF files on FIS' server
2. downloaded all these PDFs automatically (and with time gaps to avoid overloading servers)
3. scraped results data from these files according to format specifications.

This procedure was very successful, and let us quickly assemble a rather large data set with results from the 2015-2016 season. The software will be available on the GitHub repository associated with this paper, once the paper is published.

The procedure also has limitations. We exclude some events due to differences in results formats, and include only the first page of results from each file. These minor issues should not affect the results of our statistical analysis in any way. The competitions we include should be considered a random sample of competitions from the season, as the format differences are in all likelihood completely exogenous to the actual results.

>*But most of all, insights are quite limited because the authors follow a solely empiricist approach. Theoretical considerations are missing from the start, and thus the empirical findings are not very intriguing in the end:*

We concur that the original submission was less rich in theory than it was in analysis. Our aim with the study is not to make a theoretial contribution above and beyond testing implications of existing theory drawn from across the social sciences. We have however, inspired by your comments, added relevant theoretical scaffolding for our analysis. In our opinion, this move has improved the mansucript over the original version.

>* The literature review is not very extensive and only comprises empirical studies. Hypotheses refer only broadly to some of the literature or seem to be generated ex-post, instead of being deduced from any theory. Not only sociological literature is missing, but there aren't any systematic theoretical considerations, e.g. on the role of judges, on nationalism in sports, or on the specific context of ski jumping in comparison to other sports. In the end then, authors face the typical problem of such an empiricist approach: Their results - significant but weak biases, which seem to be addressed quite well by the sport's regulations - do not "speak for themselves", and thus you wonder which insight you have gained. Usually, a more thorough theoretical consideration will also lead to a more insightful interpretation of the results. Instead of treating ski jumping as just one more arbitrary case and speculate about potential threats of biases (p. 20), authors should reconstruct the social context (or field, figuration, system) of ski jumping as a specific case. Then they would be able compare their findings to other sports and reflect upon the meaning of biases in different constellations.*

Inspired by your comments, we have rethought these issues and added some material on the practice of evaluations. Here, we sketch the ski jumping field as a field in a pseudo-Bourdieusian sense and connect the national variation in judge scores to struggles of definition of what good performances are. In particular, we build on the literature on evaluation that has flowed from the seminal work of Boltanski & Thevenon, Lamont and others. We also draw upon the historical example of the Boklöv style, which at first was not taken seriously but over time has come to dominate the field.

Thank you again for your efforts! Your comments made us rethink and add components to the analysis and manuscript that in turn led to a stronger contribution to the literature.

\bigskip

END.

### 2019-07-25 revised-manuscript/manuscript.Rmd
Commit before removing home advantage

---
title: "NATIONALIST AND HOME FIELD BIAS IN SPORT PERFORMANCE EVALUATIONS"
subtitle: "An example from the ski jumping World Cup"
runhead: Judging biases in skijumping
author:
- affiliation: Department of Medical Statistics, University of Oslo
  name: Leiv Tore Salte Rønneberg
- affiliation: Department of Sociology and Human Geography, University of Oslo.
  name: Torkild Hovde Lyngstad
- affiliation: European University Institute
  name: Juho Härkönen
date: "14 september 2017"
keywords: nationalist bias, home field bias, skijumping, judges
abstract: "High stakes and prizes in international sports competitions necessitate efficiently identifying legitimate winners. Illegitimate biases of non-independent judges and home field advantages represent two major threats to the legitimacy of international competitions. Judges’ nationalities represent one source of bias, as they may prefer athletes of their own nation. Judges may also favour athletes competing within their home countries, creating another potential bias. Such biases, conscious or unconscious, may affect the outcomes of competitions. The literature offers no complete consensus on the magnitude of such biases, hence not whether they affect competition outcomes or not. In this paper, we shed light on these problems, using international ski jumping as an example. We draw on data from the FIS World Cup competitions in the 2006/2007 and 2007/2008 seasons, and use several fixed-effects based regression designs. Our results reaffirm suspicions of national bias and home field advantage in major ski jumping competitions, but also show their magnitude is too little to be of major relevance to competition outcomes. "
thanks: "The authors are grateful for financial support from the Småforsk programme at the University of Oslo, and to Marte Knutsson for research assistance in the early phase of establishing the data set.  At publication time, all code and material used to produce this paper (with the exception of the proprietary data) will be available. The authors are also grateful to Matti Nykänen for being an enduring source of inspiration in many aspects of our lives."
output:
  pdf_document:
    keep_tex: yes
    latex_engine: xelatex
    template: /home/torkildl/Dropbox/Setups/thl/rmd-xelatex-template-ms.tex
  html_document: default
bibliography: /home/torkildl/Dropbox/Literature/thl-main-library.bib
csl: /home/torkildl/Dropbox/Literature/csl-files/american-sociological-review.csl
anonymous: yes
fontsize: 12pt
geometry: margin=2cm
header-includes:
    - \usepackage{placeins}
    - \usepackage{dcolumn}
    - \usepackage{tabularx}
    - \usepackage{lscape}
    - \usepackage{caption}
fig_caption: false
fig.align: "center"
fig.width: 5
---

@@

# Read in data
jumps <- as.data.frame(read_dta("./data_prepared.dta")) %>%
    select(order(names(.))) %>%
    arrange(jumpid)
tbl_df(jumps)

@@

# Model 6: Style points model
model6 <- felm(data=d, judge ~ athome + competitionplace + judge_name + speed + length| athseason)
summary(model6)
```

@@

Many sports use qualified experts to rank and evaluate performances. For example, ski jumping competitions have a panel of five judges who independently evaluate each performance. Similar systems are used in diving, figure skating, and numerous other sports. The experts may completely or partially decide the outcome of competitions. Expert judgement is highly valued, and being a judge in sports competitions carries major prestige.

A prerogative of the system is that experts judge performances in an unbiased and neutral manner. However, there is a large potential for biases in such judgements (Meyer and Booker 1991). Experts may, for example, overestimate performances by competitors from their own country, overvalue certain stylistic elements, or unconsciously adjust evaluations in response to spectator expectations. Such biases may then be consequential in determining the outcome of competitions. Expert judgements are used well beyond the sports domain, and the relevance of studying expert judgements may inform several important societal domains. Expert judgements are widely used, for example, when evaluating applicants to universities, evaluating research grant proposals (Marsh et al. 2008), and in assessing job applicants (cf. Fasang 2006; Bursell 2012). Understanding the extent of and processes underlying such biases is important for any attempts to remove their effects.

In the present paper, we set out to study two possible types of bias in evaluations of ski jumping performances: nationalistic bias and home-field advantage bias. Nationalistic bias refers to biases in which the judges (positively) evaluate contenders from their own country, whereas home-field advantage bias arises if judges favour contenders from the country in which the contest is held. To answer our questions, we use data from the 2006-07 and 2007-08 seasons of the Federation International de Ski (FIS) World Cup series of ski jumping contests.

We contribute to the literature on nationalistic and home field advantage biases in several ways. First, by analysing both nationalistic and bias home field advantages on the same data, we can compare different aspects and sources of such bias. Second, by separately analysing the social and physical ramifications of home advantages, we can provide additional evidence on the nature of home field advantages. Third, by comparing our results to those obtained with similar data but from the markedly different context of the Olympic games, a highest-stakes-contest [@zitzewitz_nationalism_2006], we can shed light on the potential for biased evaluations in different institutional and incentive contexts.

# A REVIEW OF THE LITERATURE ON SUBJECTIVE BIASES IN SPORTS

A relatively large literature explores the existence and magnitude of subjective biases in sports evaluations. These biases may affect results or other outcomes in favor of what is in principle irrelevant aspects of competitors or other actors in the sport. Research have documented biases that promotes profits for the league owner or organization, racial and ethnic biases, nationalistic biases, as well as home advantage biases.

@@

## Nationalistic biases

@@

Vote trading in the 2002 Olympics led to a revamp of the scoring system for figure skating after a French judge admitted to being pressured to vote in favour of the Russian pair over the Canadian pair in the pairs’ competition final – effectively handing them the gold medal. Allegedly this was part of a deal with the Russian team, where they would reciprocate in a later competition in favour of the French. This scandal was examined quantitatively by Lock and Lock [-@lock_statistical_2003] using a bootstrap technique to look for inconsistencies in the scoring. Looking at the correlation between judges’ rank (of the athlete) and the final actual rank (a high correlation indicating an accurate judge), they isolated one inconsistent judge, but it was not the French one. The French judge had the best correlation of all the judges in that event; in other words, she was the most spot-on in her scoring, and no evidence of wrongdoing was found.

## Home field advantages

Home field advantage is another type of bias. Such advantages, where some competitors benefit from advantages from competing on the home field, may arise in many different ways. Several of the mechanisms that may produce home field advantages are pscyhological sources of superior performances, while others involve judges overrating home field competitors. This point illustrates that the mere presence of a biases is not necessarily illegitimate. In order to improve scoring and fairness in sports, one must understand the mechanisms behind them.

@@

### Hypotheses

We propose several hypotheses on biases in international ski jumping competitions, based on the literature reviewed above. On the basis of earlier findings, we propose a first hypothesis that there are nationalistic biases in ski jumping. Specifically, we expect there to be a positive bias in scores when judges score performances by athletes who are from the same nation as themselves.

Our second hypotheses states that the gross bias is partly explained by controlling for characteristics of the jump and athlete. The empirical implication of is that the magnitude of the bias will attenuate once all unobserved factors related to the jump are controlled.

A third hypothesis is that this bias may be partly due to cultural influences on stylistic elements, and that the strength of the bias is lower in countries with strong traditions for ski jumping and higher in countries where the sport is relatively less important.

For home field advantages, we propose that home field advantages are measurable and impacts athletes performances. We further distinguish between the social aspects of competing on the home field and the physical aspects of competing on the home field. We hypothesize that the physical home field advantage is less salient than the social home field advantage, and thus less important for performances.

@@

To test the hypotheses proposed above, we use data from the FIS World Cup competitions in ski jumping from the 2006/07 and 2007/08 seasons. The original data were downloaded from the FIS web site, coded, and collated into complete data set for analysis. Our data include detailed information on each jump in each competition, including the five individual judge scores from the jump along with the identity and nationality of each judge in each competition. We also know where the competition was held.

>>>>

---
title: "NATIONALIST AND HOME FIELD BIAS IN SPORT PERFORMANCE EVALUATIONS"
subtitle: "An example from the ski jumping World Cup"
runhead: Judging biases in skijumping
author:
- affiliation: Department of Medical Statistics, University of Oslo
  name: Leiv Tore Salte Rønneberg
- affiliation: Department of Sociology and Human Geography, University of Oslo.
  name: Torkild Hovde Lyngstad
- affiliation: European University Institute
  name: Juho Härkönen
date: "14 september 2017"
keywords: nationalist bias, home field bias, skijumping, judges
abstract: "High stakes and prizes in international sports competitions necessitate efficiently identifying legitimate winners. Illegitimate biases of non-independent judges and home field advantages represent two major threats to the legitimacy of international competitions. Judges’ nationalities represent one source of bias, as they may prefer athletes of their own nation. Judges may also favour athletes competing within their home countries, creating another potential bias. Such biases, conscious or unconscious, may affect the outcomes of competitions. The literature offers no complete consensus on the magnitude of such biases, hence not whether they affect competition outcomes or not. In this paper, we shed light on these problems, using international ski jumping as an example. We draw on data from the FIS World Cup competitions in the 2006-2008 and 2015-2016 seasons, and employ several fixed-effects based regression designs. Our results reaffirm suspicions of national bias and home field advantage in major ski jumping competitions, but also show their magnitude is too little to be of major relevance to competition outcomes. There is a trend towards smaller nationalist biases  "
thanks: "The authors are grateful for financial support from the Småforsk programme at the University of Oslo, and to Marte Knutsson for research assistance in the early phase of establishing the data set.  At publication time, all code and material used to produce this paper (with the exception of the proprietary data) will be available. The authors are also grateful to Matti Nykänen for being an enduring source of inspiration in many aspects of our lives."
output:
  pdf_document:
    keep_tex: yes
    latex_engine: xelatex
    template: /home/torkildl/Dropbox/Setups/thl/rmd-xelatex-template-ms.tex
  html_document: default
bibliography: /home/torkildl/Dropbox/Literature/thl-main-library.bib
anonymous: yes
fontsize: 12pt
geometry: margin=2cm
header-includes:
    - \usepackage{placeins}
    - \usepackage{dcolumn}
    - \usepackage{tabularx}
    - \usepackage{lscape}
    - \usepackage{caption}
fig_caption: false
fig.align: "center"
fig.width: 5
---

@@

# Read in data
jumps <- as.data.frame(read_dta(here("data_prepared.dta"))) %>%
    select(order(names(.))) %>%
    arrange(jumpid)
tbl_df(jumps)

@@

# Model 6: Style points model
model6 <- felm(data=d, judge ~ athome + competitionplace + judge_name + speed + length| athseason)
summary(model6)

#
# Season 2015-2016 analysis
#
season2016 <- data.table::fread(here("scrape/scraped-data-season-2016.csv"))

scores16 <- season2016 %>% 
    select(contains("judge"), competitorNation, country, place, speed, length, competitorName, gender) %>%
    mutate(jumpid = row_number()) %>%
    gather(key=jud, value=val, contains("judge")) %>%
    mutate(info = case_when(str_detect(jud, "Name") ~ "judge_name",
                            str_detect(jud,"Nation") ~ "judge_nation",
                            TRUE ~ "judge")) %>%
    mutate(ltr = str_sub(jud, start=-1)) %>%
    select(-jud) %>%
    group_by(jumpid, ltr) %>%
    arrange(jumpid, ltr) %>% 
    spread(key=info, value=val) %>%
    ungroup %>% 
    mutate(natbias = if_else(competitorNation==judge_nation,1,0)) %>%
    mutate_at(c("judge"), as.numeric) %>% 
    mutate(countrybias = factor(ifelse(natbias==1,as.character(judge_nation),"NON"))) %>% 
    mutate(competitionplace = str_sub(place, start=1,end=-6)) %>% 
    mutate(athome = if_else(country==competitorNation,1,0))

# NATIONAL BIAS MODELS WITH 2016 DATA
#
### Model 1b: Naïve model
model1b <- lm(data=scores16, judge ~ natbias)
summary(model1b)

### Model 2b: Introducing jump fixed effects
model2b <- felm(data = scores16, judge ~ natbias | jumpid)
summary(model2b)

### Model 3b: Introducing judge fixed effects
model3b <- felm(data = scores16, judge ~ natbias  | jumpid + judge_name)
summary(model3b)

# Model 4b: Does national bias vary among nations?
model4b <- felm(data = scores16, judge ~ relevel(countrybias,ref = "NON") | jumpid + judge_name)
summary(model4b)

# HOME FIELD BIAS MODELS WITH 2016 DATA
# 
home16 <- season2016 %>% 
    select(competitorName, place, country, competitorNation, lenpts = lengthPoints) %>% 
    mutate(competitionplace = str_sub(place, start=1,end=-6)) %>% 
    mutate(athome = if_else(country==competitorNation,1,0))

# Model 5: Length model
model5b <- felm(data=home16, lenpts ~ athome + competitionplace | competitorName)
summary(model5b)

# Model 6: Style points model
model6b <- felm(data=scores16, judge ~ athome + competitionplace + judge_name + speed + length| competitorName)
summary(model6b)

```

@@

Many sports use qualified experts to rank and evaluate performances. For example, ski jumping competitions have a panel of five judges who independently evaluate each performance. Similar systems are used in diving, figure skating, and numerous other sports. The experts may completely or partially decide the outcome of competitions.

A prerogative of the system is that experts judge performances in an unbiased and neutral manner. However, there is a large potential for biases in such judgements [@meyer_selecting_1991]. Experts may, for example, overestimate performances by competitors from their own country, overvalue certain stylistic elements, or unconsciously adjust evaluations in response to spectator expectations. Such biases may then be consequential in determining the outcome of competitions. Expert judgements are used well beyond the sports domain, and the relevance of studying expert judgements may inform several important societal domains. Expert judgements are widely used, for example, when evaluating applicants to universities, evaluating research grant proposals (Marsh et al. 2008), and in assessing job applicants (cf. @fasang_recruitment_2006; @bursell_name_2012). Understanding the extent of and processes underlying such biases is important for any attempts to remove their effects.

In the present paper, we set out to study two possible types of bias in evaluations of ski jumping performances: nationalistic bias and home-field advantage bias. Nationalistic bias refers to biases in which the judges (positively) evaluate contenders from their own country, whereas home-field advantage bias arises if judges favour contenders from the country in which the contest is held. To answer our questions, we use data from two periods, the 2006-2008 seasons and the 2015-2016 season, of the Federation International de Ski (FIS) World Cup series of ski jumping contests.

We contribute to the literature on nationalistic and home field advantage biases in several ways. First, we analyse nationalistic bias and assess to what extent such bias can be explained by unobserved characteristics of the performance. Second, we consider variations in such bias, and discuss this variation in light of sociological theory of evaluation and valuation. Third, by using our data from two time periods and comparing our results to those obtained with similar data but from the markedly different context of the Olympic games, a highest-stakes-contest [@zitzewitz_nationalism_2006], we can shed light on the role of different institutional and incentive contexts and changes over time in the magnitude of bias. Fourth, we study the social and physical ramifications of home advantages.

# THEORY AND RELEVANT LITERATURE

There is an extant scholarly literature that explores the existence and magnitude of subjective biases in sports evaluations. These biases may affect results or other outcomes in favor of what is in principle irrelevant aspects of competitors or other actors in the sport. Research have documented biases that promotes profits for the league owner or organization, racial and ethnic biases, nationalistic biases, as well as home advantage biases.

@@

@@

Vote trading in the 2002 Olympics led to a revamp of the scoring system for figure skating after a French judge admitted to being pressured to vote in favour of the Russian pair over the Canadian pair in the pairs’ competition final. This effectively handed the Russian pair the gold medal. Allegedly this was part of a deal with the Russian team, where they would reciprocate in a later competition in favour of the French. This scandal was examined quantitatively by Lock and Lock [-@lock_statistical_2003] using a bootstrap technique to look for inconsistencies in the scoring. Looking at the correlation between judges’ rank (of the athlete) and the final actual rank (a high correlation indicating an accurate judge), they isolated one inconsistent judge, but it was not the French one. The French judge had the best correlation of all the judges in that event; in other words, she was the most spot-on in her scoring, and no evidence of wrongdoing was found.

## Theorizing nationalistic biases in ski jumping

Our reading of the empirical literature on nationalistic and other evaluative biases suggests there is evidence for such biases in a variety of sports. Given its existence, what are the root causes of such bias and how do they emerge? Theoretically, we can concieve of bias in at least three different ways. We denote these three mechanisms the social psychological mechanism, the cultural legitimacy mechanism and the differential professionalization mechanism, and discuss each of them in turn below.

### The social psychological mechanism: Implicit or explicit nationalist bias

The first social psychological mechanism is the simple folk psychology theory of  judges' nationalistic impulses at the subconscious or even conscious level. They prefer their own to win, and give higher scores to those athletes. Importantly, this theoretical mechanism does not by itself predict variation in the magnitude of biases across countries. It could in princple be combined with other theoretical ideas to arrive at predictions about any such variation.

### Theory of evaluation work, as applied to nationalist biases in ski jumping

The second way of theorizing bias is to consider the role of judges and their efforts as evaluation work. Evaluative work and the justification for evaluations as a sociological study has been profoundly influences by the works of Boltanski \& Thevenon [@bolstanski_justification_1991]. A central tenet in their sociology of evaluations is the room for evaluation.

The room for evaluation decides what is relevant and good in a field. The evaluative room may be wide, and allow for many different evaluative statements, or narrow, and restrict evaluative statements. The main point is that the evaluative room allows for criticisms by field actors that in turn may affect the structure of the room for evaluation. Criticisms within the room for evaluation are recognized as forces of change in organizations and society at large. Actors cannot expect to have any criticism heard, and it must be recognized by multiple actors before it is heard. If enough actors repeat and (re)formulate  a criticism, it will potentially lead to change and a new room for evaluation.

Applied to ski jumping as a field, judges' practices for scoring represents the room for evaluation. Within this room there are likely coexisting practices that may be correlated with judges' nationalities, giving rise to national variations in judge scores. Some countries, for example, Norway and Finland, enjoy long ski jumping traditions, whereas neighbouring Sweden, an otherwise similar country, does not. Most likely the long traditions also imply that judges from these countries to a larger degree than Swedish judges will define the room for evalution. This became salient when Jan Boklöv, a Swede, pioneered the V-style in ski jumping. His style was not comme-il-faut among judges and competitors in ski jumping at the time, and thus heavily penalized by judges. Eventually he won competitions, due to the technical advantages of the V-style, and other actors started to mimic the style [@muller_performance_2008]. These changes then led to a change in the room for evaluations, where V-style was accepted and it is currently completely dominant in the field.

Mere national variation in judge scoring practices does not yield nationalistic biases in individual scores. However, when we also consider the fact that competitors are trained and socialized in their respective national ski jumping associations and competitions. This implies that young competitors share the more experienced judges' evaluative standards, and in turn this leads to nationalistic biases in judge scoring.

### Professionalisation of sports and the evolution of evaluation

The third source of national variation in ski jumping evaluations is related to the process of professionalization in sports. An essential ingredient in evaluation are standards. In sports, standards are continuously evolving. The continued professionalization of sports could would suggest a higher degree of standardization and a narrowing of the room for evaluations. Several aspects of contemporary ski jumping suggests that professionalization will limit the magnitude of nationalistic biases.

First, FIS licences judges that takes international, high-level judging assignments. Professionalization of ski jumping entails stricted quality control of judge performances and training. Currently FIS requires national skiing associations to hold at least one national-level judge seminar every year to contribute to educating new judges, and all judges are required to participate in such seminars at least once every two years.

Second, There is also a system in place for auditing judges. Chapter 10 of the FIS Rules documents explains that judges are subject to an evaluation by a FIS Sub-committee working together with a data team, and that their work may result in allotting individual judges with "points" that indicate non-fair scoring practices. If judges accumulate many points, they may not be nominated to higher-level competitions or wholly excluded from judging. Such sanctioned judges may re-qualify for judging assignments after scoring performances from television broadcasts.

Third, the competition rules also limit the influence of such biases by removing the best and worst of the five scores from the total score. This certainly limits the magnitude of any bias, but as previous research have found significant biases [@zitzewitz_nationalist_2006], it does not seem to root it completely out.

The relevance of professionalism for the *national variation* in biases emerges when we combine the idea of evaluative work with the likely different levels of professionalism among the national associations that are responsible for judge training. In countries where ski jumping is a relatively important sport, such as Finland and Norway, the bias should be less due to an assumed stronger professionalization.

Summing up, it seems meaningful to probe the existence of nationalist biases in different countries, but also study whether nationalist biased have decreased in magnitude over time, due to the general tendency towards increased  professionalization in sports.

In addition to the social psychological, cultural legitimacy, and differential professionalization mechanisms discussed above, there may well be yet other mechanisms that also contribute to nationalistic biases of this kind, for example desires on part of individual judges to grow the sport's influence in their countries and that try to boost their own competitors placements.

## The extant literature on home field advantages

Home field advantage is another type of potential bias in ski jumping. Such advantages, where some competitors benefit from advantages from competing on the home field, may arise in many different ways. Several of the mechanisms that may produce home field advantages are pscyhological sources of superior performances, while others involve judges overrating home field competitors. This point illustrates that the mere presence of a biases is not necessarily illegitimate. In order to improve scoring and fairness in sports, one must understand the mechanisms behind them.

@@

## Hypotheses

We propose several hypotheses on biases in international ski jumping competitions, based on the literature and theory reviewed above. On the basis of earlier findings, we propose a first hypothesis that there are nationalistic biases in ski jumping. Specifically, we expect there to be a positive bias in scores when judges score performances by athletes who are from the same nation as themselves.

Our second hypothesis states that the gross bias is partly explained by controlling for characteristics of the jump and athlete. The empirical implication of is that the magnitude of the bias will attenuate once all unobserved factors related to the jump are controlled.

A third hypothesis concerns variation in bias across countries. Nationalistic biases may vary across countries due to cultural-professional variations, where the strength of the bias is lower in countries with strong  jumping cultures and higher in countries where the sport is more peripheral in the larger sports field.

Our fourth hypothesis states that nationalistic biases in ski jumping evaluations decrease over time, possibly as a results of continued professionalization of the sport.

For home field advantages, we propose an additional hypothesis that home field advantages are measurable and impacts athletes performances. Finally, we distinguish between the social aspects of competing on the home field and the physical aspects of competing on the home field, and hypothesize that the physical home field advantage is less salient than the social home field advantage, and thus less important for performances.

@@

To test the hypotheses proposed above, we use data from the FIS World Cup competitions in ski jumping from the 2006-2007 and 2007-2008 seasons as well as from the 2015-2016 seasons. The original data for 2006-2007 and 2007-2008 seasons were downloaded from the FIS web site, coded, edited for errors, and collated into complete data set for analysis. For the season 2015-2016, we wrote automated web scraping software that identified relevant PDF files on the FIS servers, downloaded these files, and scraped them according to our pre-specified instructions. The software used to download and scrape the PDF files are documented in the GitHub repository associated with this article.
